Experience Report: Haskell in Computational Biology

### 2012-02-07 mrfy_experience_report.tex
a few corrections

Computational biologists write software to solve problems on large genomic or 
proteomic data sets. Traditionally, these software programs fall into two 
categories: those with high-performance computing needs are typically 
implemented in C++, while the remainder are typically implemented in Perl, 
Python, Ruby, or R. Recently, we have developed a program in Haskell which 
implements a novel algorithm for solving the remote homology detection problem. 
During this process, we have wondered: would the algorithm be easily expressed 
in Haskell? Would we be able to achieve acceptable runtime performance? How 
would our experience compare with our past experiences developing and 
maintaining computational biology software in C++, Ruby, and Python? What would 
be our greatest challenges? In this paper, we briefly explain the remote 
homology detection problem and our algorithm, and then attempt to answer the 
above questions.

@@

All three authors are members of the Tufts University Department of Computer 
Science. Noah Daniels has taken a graduate seminar in advanced functional 
programming, which included some Haskell, and spent ten years in industry as a 
professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
has taken a programming language heavily focused on functional programming, but 
with no Haskell. Norman Ramsey is the local expert, but he wrote no code for 
this project.

@@

The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
relations that seek to maximize the probability of an observed sequence (in 
this case, of amino acid ``letters'') begin emitted by a particular 
finite-state automaton (the hidden Markov model). The particular states of this 
finite-state automaton map to the recurrence relations, and in this instance 
derive from what is known in the computational biology community as a 
\textit{profile} hidden Markov model. The Viterbi algorithm in this instance 
recurses on ``match,'' ``insertion,'' and ``deletion'' states. To be efficient, 
the Viterbi algorithm relies on dynamic programming, or memoization, and the 
resulting asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence 
of distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence. In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements.

>>>>

Computational biologists write software to solve problems on large genomic or 
proteomic data sets. Traditionally, these software programs fall into two 
categories: those with high-performance computing needs are typically 
implemented in C++, while the remainder are typically implemented in Perl, 
Python, Ruby, or R. Recently, we have developed a program in Haskell which 
implements a novel algorithm for solving the remote homology detection problem. 
While designing the algorithm, we wondered: would the algorithm be easily 
expressed in Haskell? Would we be able to achieve acceptable runtime 
performance? How would our experience compare with our past experiences 
developing and maintaining computational biology software in C++, Ruby, and 
Python? What would be our greatest challenges? In this paper, we briefly 
explain the remote homology detection problem and our algorithm, and then 
attempt to answer the above questions.

@@

All three authors are members of the Tufts University Department of Computer 
Science. Noah Daniels has taken a graduate seminar in advanced functional 
programming, which included some Haskell, and spent ten years in industry as a 
professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
has taken a programming languages course heavily focused on functional 
programming, but with no Haskell. Norman Ramsey is the local expert, but he 
wrote no code for this project.

@@

The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
relations that seek to maximize the probability of an observed sequence (in 
this case, of amino acid ``letters'') being emitted by a particular 
finite-state automaton (the hidden Markov model). The particular states of this 
finite-state automaton map to the recurrence relations, and in this instance 
derive from what is known in the computational biology community as a 
\textit{profile} hidden Markov model. The Viterbi algorithm in this instance 
recurses on ``match,'' ``insertion,'' and ``deletion'' states. To be efficient, 
the Viterbi algorithm relies on dynamic programming, or memoization, and the 
resulting asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence 
of distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence. In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements.
